\documentclass{amsart}
\usepackage[margin=1.2in]{geometry}
\usepackage{fancyhdr}
\usepackage{graphicx,hyperref,xcolor} 
% Set up the fancy header and footer
\pagestyle{fancy}
\fancyhead{}  % Clear all headers and footers
\fancyfoot[C]{\thepage} 

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={DD-MED},
    pdfpagemode=FullScreen,
    }

\title{CA STV Modeling}
\author{Matthew Jones, Hane Lee, Jay Lee ?, Max Springer with support from Christopher Donnay, Moon Duchin, and Peter Rock.}
\date{\today}


\begin{document}

\maketitle

\section{How it works}
Single transferable vote, or STV, is a ranked-choice voting system that can be used to elect 
one or more seats.\footnote{In the case of electing a single seat, STV is often called 
instant-runoff voting (IRV).} In STV, a threshold of first-place votes is required to be elected.
If no candidate meets the threshold, the candidate with the fewest first-place votes is eliminated,
and their votes are transferred to the next candidate on those ballots.
This is repeated until a candidate reaches the threshold; the candidate is elected, and
their votes in surplus of the threshold are transferred to the next candidate on those ballots.
This is repeated until all seats have been filled.

Here is a simple example of how STV works. \todo{Add example}

In our example above, there were a few algorithmic decisions that had to be made. 
The first is the threshold. 
Above we used the \emph{Droop quota}, the most common threshold used in STV elections.
Let $m$ denote the number of seats to be filled ($m$ stands for \emph{magnitude}).
The Droop quota is calculated as
\[\left\lfloor\frac{\text{\# of votes}}{m+1} + 1\right\rfloor,\]
where $\lfloor \cdot \rfloor$ denotes rounding down to the nearest integer.
One way to interpret the Droop quota is that it is the minimum number of votes such 
that no more than $m$ candidates can achieve the quota.
There are other thresholds that can be used, including some which updated dynamically
as the tabulation process progresses.\todo{Cite}
For the remainder of this report, we will use the Droop quota.

Another important decision is the method for transferring votes. 
One option is called \emph{fractional transfer}. In fractional transfer, once a candidate is elected,
any ballot with that candidate in first place is discounted by a factor of 
\[\frac{\text{\# of votes}- \text{threshold}}{\text{\# of votes}},\]
and the vote transfers to the next candidate on the ballot.
This is how we transferred votes in our example above.
In Cambridge, Massachusetts, they use a method called \emph{random transfer}.\todo{Cite}
In random transfer, once a candidate is elected, they randomly select 
\[\text{\# of votes}- \text{threshold}\]
ballots to keep, and the votes are transferred to the next candidate on those ballots.
The unselected ballots are no longer used in the tabulation process.

There are other transfer methods? including \todo{cite} 

Finally, there are two edge cases that need to be handled.
The first is the method for handling  multiple candidates 
crossing the threshold in a single round. Either elect all of them in that round,
or elect the one with the most votes, and proceed to another round of tabulation.
This decision affects the order in which candidates are elected and which can even 
alter the winner set.
The second consideration is a method for handling ties. Ties can occur whenever two or more
candidates have the same number of first-place votes.\todo{Cite, https://www.portland.gov/sites/default/files/2024/Portland-District-1-Certified-Abstract-Nov-2024.pdf}\footnote{This is not an edge case; ties occur 
in real-world STV elections. See Round 8 in District 1 of the Portland, OR city council election in 2024.}
One method is to just randomly select one of the tied candidates; another is to use the 
first-place vote count from the first round of tabulation.  

Taken together, these algorithmic decisions make up an STV election.

\section{Background and prior use}
International and US 

\section{Pros and Cons}
Candidate-centeredness, district size limitations, advocacy in US/FairVote, implications for party behavior
\section{Specific concerns}
Methods of translating votes to seats, vote leakage and truncation
CA Single-subject rule concerns

\section{Modeling STV in CA}
CA Model – size of districts, ballot design Implications for descriptive and ideological representation in CA
\end{document}